\documentclass[letterpaper,12pt]{article}
\setlength{\headheight}{14.49998pt}
\usepackage{fancyhdr}
\usepackage{lipsum,graphicx}
\usepackage{amsmath, amsfonts, amssymb, ragged2e}
\usepackage{amsthm}
\usepackage{bookmark}
\usepackage{bbm}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}

\title{Summary of ELEC 433}
\author{Tom Wang}
\date{Spring, 2024}

\fancypagestyle{plain}{
    \fancyhf{}
    \fancyhead[L]{Tom Wang}
    \fancyhead[R]{\thepage}
}

\begin{document}

\maketitle
\section{Introduction to Hamming Codes}
ECC solves the problem of errors in transmission. The idea is to add redundancy to the message so that the receiver can detect and correct errors.

Assume that we have a message of length $k$, then we encode it to a codeword $C$ of length $n$. If the codeword is decrypted, then we call it $\tilde{C}$.

\subsection{Things we care about}
\begin{enumerate}
    \item \textbf{Error Detection}: We want to be able to detect if there are errors in the transmission.
    \item \textbf{Error Correction}: We want to be able to correct the errors in the transmission.
    \item \textbf{Efficiency}: We want to be able to minimize the number of bits we need to add to the message. So minimize the overhead. ($\frac{k}{n}$ should be as large as possible)
    \item \textbf{Complexity}: We want to be able to encode and decode the message in a reasonable amount of time.
    \item \textbf{Robustness}: We want to be able to correct as many errors as possible.
    \item \textbf{Flexibility}: We want to be able to correct different types of errors.
\end{enumerate}
\subsection{Definitions}
First of all, we need to define a code:

\begin{definition}
    A code $C$ of length $n$ over an alphabet $\Sigma$ is a subset of $\Sigma^n$. So $C\subseteq \Sigma^n$. The elements $c\in C$ are called codewords.
\end{definition}

There are two types of errors:\begin{itemize}
    \item \textbf{Erasure}: We know which bit is erased, but we don't know what it is. So we can \textbf{detect} and \textbf{correct} erasures.
    \item \textbf{Error}: We don't know which bit is wrong, but we know that there is at least one bit that is wrong. So we can \textbf{detect} errors, but we can't for sure \textbf{correct} them.
\end{itemize}

Also, we define all possible codewords to be $C_i$ where they all belong to the same set $\mathcal{C}$. So $C_i\in \mathcal{C}$ for all possible $i$.
\subsection{Hamming Codes}
Hamming codes are linear block codes. They are named after Richard Hamming. They are single-error correcting and double-error detecting. They are also systematic, which means that the message is part of the codeword.

Hamming code encodes a message of length 4 to a codeword of length 7. So $k=4$ and $n=7$. Also, it encodes only binary messages, so $\Sigma = \{0,1\}$.

\[(x_1,x_2,x_3,x_4)\implies (x_1,x_2,x_3,x_4,x_2+x_3+x_4,x_1+x_3+x_4,x_1+x_2+x_4)\]

Note that they are binary additions. So $1+1=0$. It is basically XOR.

Some definitions:\begin{itemize}
    \item \textbf{Hamming distance}: Hamming distance between $x,y\in \Sigma^n$ is the number of positions where $x$ and $y$ differ. It is denoted by $d(x,y)=\sum_{i=1}^{n}\mathbbm{1}\{x_i\neq y_i\}$. So adding each bit that is different.
    \item \textbf{Relative distance}: counts for the length of the codeword. So $\frac{d(x,y)}{n}$.
    \item \textbf{Minimum distance}: The minimum distance between any two codewords. So $d_{min} = \min d(x,y)$, for which ${x,y\in C, x\neq y}$. It measures the robustness of the code. The larger the minimum distance, the more errors it can correct.
    \item \textbf{Hamming weight}: The number of non-zero elements in a vector. So $w(x)=\sum_{i=1}^{n}\mathbbm{1}\{x_i\neq 0\}$. It is the number of 1s in the vector.
\end{itemize}

\subsection{Hamming Code Properties}
\begin{theorem}
    A code with a minimum distance of $d$ can:\begin{itemize}
        \item Detect $d-1$ errors.
        \item Correct $\lfloor \frac{d-1}{2}\rfloor$ errors.
        \item Detect $d-1$ erasures.
        \item Correct $d-1$ erasures.
    \end{itemize}
\end{theorem}

\subsection{Algorithm for Hamming Code}
\subsubsection{Error Detection}
If given a codeword $\tilde{C}$ that we assume the error is within $d-1$ away from one point, we can calculate the syndrome $S$ by $S = \tilde{C}H^T$. If $S=0$, then there is no error. If $S\neq 0$, then there is an error.

\subsubsection{Error Correction}
For a codeword $\tilde{C}$, with an error smaller than or equal to $\lfloor \frac{d-1}{2}\rfloor$, we can recover the original codeword by finding the closest codeword to $\tilde{C}$. The distance will be within $\lfloor \frac{d-1}{2}\rfloor$.

Note that it is not possible to have two $C$ that are within $\lfloor \frac{d-1}{2}\rfloor$ distance from $\tilde{C}$. Otherwise, the distance between the two $C$ will be smaller than $d$. Here is the proof:
\begin{proof}

    $\lfloor \frac{d-1}{2}\rfloor\times 2\le d-1<d$. So circle of radius $\lfloor \frac{d-1}{2}\rfloor$ around $\tilde{C}$ does not contain any other codewords. So there is only one codeword that is within $\lfloor \frac{d-1}{2}\rfloor$ distance from $\tilde{C}$.
\end{proof}

To mathematically express this:

If $\tilde{C}$ has error $\le \lfloor \frac{d-1}{2}\rfloor$, then return $C\in\mathcal{C}$ such that $d(\tilde{C},C)\le \lfloor \frac{d-1}{2}\rfloor$.

\subsubsection{Erasure Correction}
If we have a codeword $\tilde{C}$ with $e$ erasures, where $e\le d-1$, then we can recover the original codeword by finding the closest codeword to $\tilde{C}$. The distance will be within $e$.

\subsection{More definitions}
\begin{definition}
    The \textbf{message length/dimension} of $\mathcal{C}$ over an alphabet $\Sigma$ is the length of the message. So $k=log_{|\sigma|}{|\mathcal{C}|}$.
\end{definition}
(The log base is the size of the alphabet set, and the log is the size of the code set)

\begin{definition}
    The \textbf{rate} of $\mathcal{C}$ is the ratio of the message length to the codeword length. So $R=\frac{k}{n}$.
\end{definition}
Note the rate is between 0 and 1. \begin{itemize}
    \item The closer it is to 1, the more efficient the code is.
    \item If $R=1$, then the code is not redundant. So it is not an error-correcting code.
    \item  The closer it is to 0, the more redundant the code is.
    \item If $R=0$, then the code is not efficient. So it is not an error-correcting code.
    \item So we need to make trade-offs between efficiency and robustness.
\end{itemize}

We could find that $R$ is a function: $R(d,n,k,\Sigma)$. So we can find the maximum rate and minimum rate for a given $d,n,k,\Sigma$. Basically, R has two bounds:
\[g(d,n,k,\Sigma)\le R(d,n,k,\Sigma) \le f(d,n,k,\Sigma)\]
\subsection{Hamming Bound}
Hamming bound is the upper bound of the rate of a code. It answers the question: How many Hamming balls can we put in $\Sigma^n$ such that they are disjoint for a fixed $d$?

The question implies that the ball radius is $\lfloor \frac{d-1}{2}\rfloor$. With the given information we can find that \begin{itemize}
    \item We will have $|\mathcal{C}|$ disjoint balls of radius $\lfloor \frac{d-1}{2}\rfloor$.
    \item Vol$(\Sigma^n)=|\Sigma|^n$.
    \item $|\mathcal{C}|\le \frac{|\Sigma|^n}{\text{Vol(Hamming ball)}}$.
\end{itemize}
Note that we are dealing with discrete points. So we need to use the floor function which actually just counts for the points within the ball.


\subsubsection{Proof of Hamming Bound}
\begin{proof}

    The Hamming ball in $\Sigma^n$ of radius $e$ about $x\in \Sigma^n$ is the set of all points in $\Sigma^n$ that are within $e$ distance from $x$.

    So $B_{\Sigma^n} (x,e)=\{y\in \Sigma^n: d(x,y)\le e\}$ is the set of the discrete points that are within $e$ distance from $x$, including $x$ itself.

    Now we can define the volume of the ball. The volume of the ball is the number of points in the ball. So $V_{\Sigma^n}=|B_{\Sigma^n} (x,e)|$ which is the size of the ballpoint set.

    To make the calculation easier, we let $q = \Sigma^n$. \[
        \text{Vol}_q(e,n)= 1+ \binom{n}{1}(q-1)+\binom{n}{2}(q-1)^2+\ldots+\binom{n}{e}(q-1)^e=\sum_{i=0}^{e}\binom{n}{i}(q-1)^i
    \]

    So since Hamming balls cannot overlap, we have: \begin{align*}
        |\mathcal{C}|                                   & \le \frac{q^n}{\text{Vol}_q(e,n)}         \\
        |\mathcal{C}|\times \text{Vol}_q(e,n)           & \le q^n                                   \\
        \log_q(|\mathcal{C}|)+\log_q(\text{Vol}_q(e,n)) & \le \log_q(q^n)                           \\
        \log_q(|\mathcal{C}|)+\log_q(\text{Vol}_q(e,n)) & \le n                                     \\
        \log_q(|\mathcal{C}|)                           & \le n-\log_q(\text{Vol}_q(e,n))           \\
        \textbf{Rate}= \frac{\log_q(|\mathcal{C}|)}{n}  & \le 1-\frac{\log_q(\text{Vol}_q(e,n))}{n}
    \end{align*}
    The rate here is the maximum rate. So we have the upper bound of the rate of a code. So we have the \textbf{Hamming bound}:
\end{proof}
\subsubsection{Analysis of Hamming Bound for binary codes}
Now that we have all the tools for binary Hamming codes:\begin{itemize}
    \item $q=2$. So $|\Sigma|=2$, which is the alphabet size.
    \item $n=7$. So $n$ is the length of the codeword.
    \item $k=4$. So $k$ is the length of the message.
    \item $d=3$. So $d$ is the minimum distance. Proof of this will be in the next section.
\end{itemize}
Using the above information, we find that the Hamming bound is:\[
    R = 1-\frac{\log_2(\text{Vol}_2(\lfloor \frac{3-1}{2}\rfloor, 7))}{7}= 1-\frac{\log_2(1+7)}{7}=\frac{4}{7}
\]
But we also notice that the Hamming code has a rate of $\frac{k}{n}=\frac{4}{7}$. So the Hamming bound is tight. So the Hamming code is the best/optimal code for this case. 

\subsection{Linear Algebra View of Hamming Codes}
If we view the encoding process as a matrix multiplication it will be like this:
\[
    \begin{bmatrix}
        1 & 0 & 0 & 0 \\
        0 & 1 & 0 & 0 \\
        0 & 0 & 1 & 0 \\
        0 & 0 & 0 & 1 \\
        0 & 1 & 1 & 1 \\
        1 & 0 & 1 & 1 \\
        1 & 1 & 0 & 1
    \end{bmatrix}
    \begin{bmatrix}
        x_1 \\
        x_2 \\
        x_3 \\
        x_4
    \end{bmatrix}
    =
    \begin{bmatrix}
        x_1 \\
        x_2 \\
        x_3 \\
        x_4 \\
        x_2+x_3+x_4 \\
        x_1+x_3+x_4 \\
        x_1+x_2+x_4
    \end{bmatrix}
    \]
    
We can write it as $G\cdot X=\mathcal{C}$ where $G$ is the generator matrix. So $G$ is a $n\times k$ matrix. $X$ is the message vector. $\mathcal{C}$ is the codeword vector. $\mathcal{C}=\{G\cdot X (\mod 2): \forall x \in X, x\in \{0,1\}^4\}$

There are three properties with it:\begin{enumerate}
    \item $\mathcal{C}$ is closed under addition. So $\forall C_1,C_2\in \mathcal{C}, C_1+C_2\in \mathcal{C}$.\begin{proof}
        \[
            C_1+C_2 = G\cdot X_1+G\cdot X_2 = G\cdot (X_1+X_2)
        \]
        Since $X_1+X_2\in \{0,1\}^4$, so $C_1+C_2\in \mathcal{C}$. Proved.
    \end{proof}
    \item $\mathcal{C}$ is a linear subspace of $\{0,1\}^7$ of dimension 4. \begin{proof}
        $\mathcal{C}$ is all the linear combinations of the columns of $G$. So $\mathcal{C}$ is a linear subspace of $\{0,1\}^7$ of dimension 4. Proved.
    \end{proof}
    \item The distance of $\mathcal{C}$ is the same as the minimum weight of the non-zero codewords in $\mathcal{C}$.\begin{proof}
        Recall that $wt(x)=\sum_{i=1}^{n}\mathbbm{1}\{x_i\neq 0\}$. So $wt(x)$ is the number of 1s in the vector. 
        
        $\forall C_1,C_2\in \mathcal{C}, C_1\neq C_2$, then $wt(C_1+C_2)=wt(C_1 \oplus C_2)$,which is just the number of different bits between $C_1$ and $C_2$. So $d(C_1,C_2)=wt(C_1 \oplus C_2)$. 

        Thus if we go through all the possibility space, we can claim that $d_{min}=\min_{C_1, C_2\in \mathcal{C}, C_1\neq C_2}d(C_1, C_2)=\min_{C\in \mathcal{C}, C\neq 0}wt(C)$. Proved.
    \end{proof}
\end{enumerate}

\subsubsection{Parity Check Matrix}
We can also define a parity check matrix $H$ for the Hamming code. It is a $n\times (n-k)$ matrix. So $H\cdot C^T=0$ where $C$ is the codeword vector. So $H$ is the matrix that can detect errors.
\[
    \begin{bmatrix}
        0 & 1 & 1 & 1 & 1 & 0 & 0 \\
        1 & 0 & 1 & 1 & 0 & 1 & 0 \\
        1 & 1 & 0 & 1 & 0 & 0 & 1
    \end{bmatrix}
    \begin{bmatrix}
        x_1 \\
        x_2 \\
        x_3 \\
        x_4 \\
        c_5 = x_2+x_3+x_4 \\
        c_6 = x_1+x_3+x_4 \\
        c_7 = x_1+x_2+x_4
    \end{bmatrix}
    =
    \begin{bmatrix}
        0 \\
        0 \\
        0
    \end{bmatrix}
\]

Note that $H$ is just letting the encoded codeword to XOR with its decomposition. It should be 0 if there is no error. So if there is an error, then $H\cdot C^T\neq 0$.

\subsubsection{Proof of Minimum Distance}
With the help of the parity check matrix, we can prove that the minimum distance of the Hamming code is 3.\begin{proof}
    Proof by contradiction: Suppose that $\exists C \in \mathcal{C}$ such that $wt(C)=1$ or $wt(C)=2$. 

    We can list out all the possibilities of C and find out that: $H\cdot C^T\neq 0$. So $C$ is not a codeword. Contradiction. So $\forall C\in \mathcal{C}, wt(C)\ge 3$. 
    
    Now just need to find a codeword with weight 3. We can find that $C=(1,1,1,0,0,0,0)$ is a codeword. With the property that we just proved: 
    
    \textit{The distance of $\mathcal{C}$ is the same as the minimum weight of the non-zero codewords in $\mathcal{C}$.} 
    
    So $d_{min}=3$. Proved.
\end{proof}

\subsection{Decoding Hamming Codes}
Assume that we have a codeword $\tilde{C}$ that is transmitted. We want to find the closest codeword to $\tilde{C}$.\begin{itemize}
    \item If $\tilde{C}$ has no error, then we can just return $\tilde{C}$.
    \item If $\tilde{C}$ has one error, $\tilde{C}=C\oplus Z$ where $Z$ is the noise such that $wt(Z)=1$. \begin{align*}
        H\cdot \tilde{C}^T & = H\cdot (C\oplus Z)^T \\
                           & = H\cdot C^T\oplus H\cdot Z^T &\text{Addition is commutative}\\
                           & = H\cdot Z^T &\text{Since $H\cdot C^T=0$}
    \end{align*}
    From there, we can know which bit is wrong now, and moreover, we can correct it.\[
        C = \tilde{C}\oplus Z
    \] XOR can be reversed by XOR again. So we can correct the error.
\end{itemize}
\end{document}