\documentclass[letterpaper,12pt]{article}
\setlength{\headheight}{14.49998pt}
\usepackage{fancyhdr}
\usepackage{lipsum,graphicx}
\usepackage{amsmath, amsfonts, amssymb, ragged2e}
\usepackage{amsthm}
\usepackage{bookmark}
\usepackage{bbm}
\usepackage{listings}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}

\title{Summary of CPSC 221}
\author{Tom Wang}
\date{Spring, 2024}

\fancypagestyle{plain}{
    \fancyhf{}
    \fancyhead[L]{Tom Wang}
    \fancyhead[R]{\thepage}
}


\begin{document}
\maketitle
\section{Complexity}
A ``good'' algorithm should:\begin{itemize}
    \item Be correct (or at least mostly correct)
    \item finish in a reasonable amount of time (time complexity)
    \item uses a reasonable amount of memory (space complexity)
\end{itemize}

\subsection{Time Complexity}
Running time is expressed as a function $T(n):\mathbb{Z}^0 \to \mathbb{R}^0$, where $n$ is the size of the input. We are interested in the asymptotic behavior of $T(n)$ as $n\to \infty$. We use big-O notation to express this.
\subsubsection{Asymptotic Notation}
\begin{definition}

    $T(n)\in O(f(n))$ if there exists $c>0$ and $n_0>0$ such that $T(n)\leq cf(n)$ for all $n\geq n_0$.

    $T(n)\in \Omega(f(n))$ if there exists $c>0$ and $n_0>0$ such that $T(n)\geq cf(n)$ for all $n\geq n_0$.

    $T(n)\in \Theta(f(n))$ if $T(n)\in O(f(n))$ and $T(n)\in \Omega(f(n))$.

    $T(n)\in o(f(n))$ if for all $c>0$, there exists $n_0>0$ such that $T(n)\leq cf(n)$ for all $n\geq n_0$.

    $T(n)\in \omega(f(n))$ if for all $c>0$, there exists $n_0>0$ such that $T(n)\geq cf(n)$ for all $n\geq n_0$.
\end{definition}

The reason we use big-O notation and omit the details is that it is easy to work with. For example, if $T(n)\in O(f(n))$ and $S(n)\in O(g(n))$, then $T(n)+S(n)\in O(f(n)+g(n))$ and $T(n)S(n)\in O(f(n)g(n))$.

We only care about the dominant term in the function (When n gets really large). We also ignore constant factors because we are only interested in the asymptotic behavior.

\subsubsection{Analyze loops}
\begin{lstlisting}[language=C++]
bool hasDuplicate(int arr[], int size){
    for(int i=0; i<size-1; i++){
        for(int j=+1; j<size; j++){
            if(arr[i]==arr[j]){
                return true;
            }
        }
    }
    return false;
}
\end{lstlisting}

Now we analyze the running time of this algorithm. Basically, counts for the worst-case scenario.

It is easy to tell that the outer loop runs $n-1$ times, thus $O(n)$.

The inner loop runs from $n-1$ to $1$ times if we run the loop. However, we can use the fact that the sum of the first $n$ integers is $\frac{n(n-1)}{2}$. So the inner loop runs $\frac{n(n-1)}{2}$ times, thus $O(n^2)$. This is the overall complexity of the nested loop.

Give more examples:\begin{lstlisting}[language=C++]
void candyapple(int n) {
    for (int i = 1; i < n; i *= 3)
    cout << "iteration: " << i << endl;
}
void caramelcorn(int n) {
    for (int i = 0; i * i < 6 * n; i++)
    cout << "iteration: " << i << endl;
}         
\end{lstlisting}
If the loop variable is multiplied by a constant, then the loop runs in $O(\log n)$ time. If the loop variable is squared, then the loop runs in $O(\sqrt{n})$ time.

So candyapple runs in $O(\log n)$ time and caramelcorn runs in $O(\sqrt{n})$ time.
\begin{proof}

    in candyapple, $i$ is multiplied by 3 each time, so $i=3^k$ after $k$ iterations. So $3^k=n$, so $k=\log_3 n$. So the loop runs $\log_3 n$ times, thus $O(\log n)$.

    in caramelcorn, $i^2=6n$, so $i=\sqrt{6n}$. So the loop runs $\sqrt{6n}$ times, thus $O(\sqrt{n})$.
\end{proof}

lastly, we can analyze the following code:\begin{lstlisting}[language=C++]
int i, j;
for (i = 1; i < 9*n; i = i*2) {
    for (j = n*n; j > 0; j--) {
        ...
    }
}
\end{lstlisting}
The outer loop runs $\log_2 9n$ times, thus $O(\log n)$. The inner loop runs $n^2$ times, thus $O(n^2)$. So the overall complexity is $O(n^2\log n)$.

\section{Correctness}
Former methods to argue Correctness:
\subsection{Loop Invariant}
\begin{definition}
    \begin{itemize}
        \item A loop invariant is a property that holds before and after each iteration of a loop.
        \item Within a loop body, these properties may be violated briefly but must be fixed for the next iteration.
        \item A loop invariant proof must use features of the code being analyzed to demonstrate that any violations of loop invariants are fixed.
    \end{itemize}
\end{definition}
\subsubsection{Proving a loop invariant}
Generally, we use induction to prove it:\begin{enumerate}
    \item Find an induction variable or property.
    \item Base case: show that the loop invariant holds before the first iteration.
    \item Inductive hypothesis: assume that the loop invariant holds before the $k$th iteration. $k$ can be unspecified.
    \item Inductive step: show that the loop invariant holds for the $(k+1)$th iteration.
    \item Termination: show that the loop invariant holds after the last iteration.
\end{enumerate}
\subsection{Selection Sort}
\begin{lstlisting}[language=C++]
void selectionSort(Vector<int> & arr) {
    for(int i=0; i<arr.size(); i++){
        int min = indexOfMin(arr, i);
        swap(arr[i], arr[min]);
    }
}
\end{lstlisting}
This algorithm finds the smallest element in the array and swap it with the first element. Then it finds the second smallest element and swap it with the second element. And so on.

For each iteration, we need to compare $n-i$ elements to find the smallest element. So a total of $\sum_{i=0}^{n-1}n-i=\frac{n(n-1)}{2}$ comparisons. Swap is just a function to perform each iteration, which could be seen as $O(n)$. So the running time is $O(n^2)$.

Note that the running time is independent of the input. So it is a \textbf{deterministic} algorithm. Running time is the same for best, average, and worst case.

Space complexity is $O(1)$ because we only need to store the array and use a few variables on the stack to keep track of the indices.
\subsubsection{Invariant proof of selection sort}
\begin{proof}
    Claim that the loop invariant is that the first $i$ elements are sorted. All elements before $i$ are smaller than or equal to all elements after $i$.

    Base case: before the first iteration, $i=0$, so the first $i$ elements are empty, so the invariant holds. 

    Inductive hypothesis: assume that the invariant holds before the $k$th iteration.

    Inductive step: after the $k$th iteration, the smallest element in the array is swapped with the $k$th element. So the first $k$ elements are sorted. All elements before $k$ are smaller than or equal to all elements after $k$.

    Termination: after the last iteration, $i=n$, so the first $i$ elements are the whole array, which is sorted. All elements before $i$ are smaller than or equal to all elements after $i$.
\end{proof}
\subsection{Insertion Sort}
\begin{lstlisting}[language=C++]
void slide(Vector<int> & arr, int p) {
    //need arr is in ascending order up to p exclusive
    int temp = arr[p];
    int j = p;
    while(j>0 && arr[j-1]>temp){
        arr[j] = arr[j-1];
        j--;
    }
    arr[j] = temp;
}
void insertionSort(Vector<int> & arr) {
    for(int i=1; i<arr.size(); i++){
        slide(arr, i);
    }
}
\end{lstlisting}
\subsubsection{Slide}
Slide checks the pth element and moves to the front of the array if it is smaller than the previous elements. It is a $O(n)$ operation (worst case) and $\Omega(1)$ (best case) space complexity.

loop invariant:  \begin{itemize}
    \item the first $p$ elements are sorted.
    \item $arr[0:j-1]\cup arr[j+1:p] \cup temp$ is the same value \textbf{set} as $arr[0:p]$.
\end{itemize}
\subsubsection{Insertion Sort}
Insertion sort is a $O(n^2)$ algorithm since it calls slide $n-1$ times where slide is $O(n)$.

Best case is a sorted array, so the running time is $O(n)$.

Worst case is a reverse sorted array, so the running time is $O(n^2)$.

If random data is given, the running time is closer to the worst case than the best case. Around $\frac{n(n-1)}{4}$ comparisons and $\frac{n(n-1)}{4}$ swaps.

Insertion sort is good for small data sets and nearly sorted data. Also, it barely uses any space (memory), so it is good for embedded systems.

\subsubsection{correctness proof}
\begin{proof}
    \textbf{Base case:} before the first iteration, $i=1$, so the first $i$ elements are sorted. All elements before $i$ are smaller than or equal to all elements after $i$.

    \textbf{Inductive hypothesis:} assume that the invariant holds before the $k$th iteration.

    \textbf{Inductive step:} after the $k$th iteration, the $k$th element is inserted into the first $k$ elements. So the first $k+1$ elements are sorted. All elements before $k+1$ are smaller than or equal to all elements after $k+1$.

    \textbf{Termination:} after the last iteration, $i=n$, so the first $i$ elements are the whole array, which is sorted. All elements before $i$ are smaller than or equal to all elements after $i$.

    All of the above is true because of the loop invariant of slide.
\end{proof}

\section{Memory}
4 area of memory: \begin{enumerate}
    \item Code: the compiled machine code of the instructions.
    \item Stack: local variables, function parameters, and return addresses. Automatically allocated and deallocated. (Usually about few MB)
    \item Heap: dynamically allocated memory. Must be manually allocated and deallocated. (Usually about few GB)
    \item NULL: the area of memory that is not used. Address 0x0.
\end{enumerate}
\subsection{Call by value and call by reference}
\begin{lstlisting}[language=C++]
void swap_wrong(int a, int b){
    int temp = a;
    a = b;
    b = temp;
}
void swap_true(int & a, int & b){
    int temp = a;
    a = b;
    b = temp;
}
int main(){
    int x = 3;
    int y = 5;
    swap(x, y);
    cout << x << " " << y << endl;
}  
\end{lstlisting}
The output of the above code is 3 5. This is because the swap function is called by value, so the original x and y are not changed.

If we change the swap function to swap\_true, then the output will be 5 3. This is because the swap function is called by reference, so the original x and y are changed.

\subsection{Pointers}
Pointers are variables that store the address of another variable. It is a powerful tool but also dangerous. It is powerful because it can access the memory directly. It is dangerous because it can access the memory directly.

\begin{lstlisting}[language=C++]
int main(){
    int x = 3;
    int * p = &x;
    cout << *p << endl;
    *p = 5;
    cout << x << endl;
}
\end{lstlisting}
The output of the above code is 3 5. This is because the pointer p is pointing to the address of x. So *p is the value of x. So *p = 5 is the same as x = 5.

Note that the type of p is int*, which means it is a pointer to an integer. The type of *p is int, which means it is the value of the integer that p is pointing to.

\subsubsection{new and delete}
\begin{lstlisting}[language=C++]
int main(){
    int * p = new int;
    *p = 5;
    cout << *p << endl;
    delete p;
}
\end{lstlisting}
The above code is an example of dynamic memory allocation. The new operator allocates memory for an integer and returns a pointer to the memory. The delete operator deallocates the memory. It will give memory space in the heap instead of the stack. Thus it is a good way to avoid stack overflow if we want some huge data structure.

\subsubsection{Memory leak}
\begin{lstlisting}[language=C++]
int main(){
    int * p = new int;
    p = new int;
    delete p;
}
\end{lstlisting}
The above code is an example of memory leak. The first new operator allocates memory for an integer and returns a pointer to the memory. The second new operator allocates memory for another integer and returns a pointer to the memory. The delete operator deallocates the memory for the second integer. However, the memory for the first integer is not deallocated. This is a memory leak.




















\end{document}